---
title: "Getting Started with mlbplotR"
description: "A brief introduction to the ggplot2 extension."
author: "Camden Kay"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mlbplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 600
)


# Preset dataframes
library(dplyr)

filtered_df <- tibble::tribble(
  ~Team, ~ERA, ~FIP, ~HR,
  "ARI",  4.3, 4.25,  19,
  "ATL", 3.04, 3.31,  15,
  "BAL", 3.62, 4.62,  30,
  "BOS", 3.75, 2.79,  15,
  "CHC", 4.77, 4.89,  31,
  "CHW", 2.37, 2.65,  13,
  "CIN", 3.37, 3.97,  17,
  "CLE", 3.12, 4.05,  14,
  "COL", 4.33, 4.63,  20,
  "DET", 3.71, 4.71,  24,
  "HOU", 3.14, 4.01,  12,
  "KCR", 4.77,  5.6,  22,
  "LAA", 3.18, 3.52,  15,
  "LAD", 2.47, 3.16,  19,
  "MIA", 2.64, 2.55,   6,
  "MIL", 2.43, 1.63,   7,
  "MIN", 3.72, 4.27,  17,
  "NYM", 3.02, 3.49,  17,
  "NYY", 3.23, 2.92,  24,
  "OAK", 3.15, 3.34,  15,
  "PHI", 2.78, 2.59,  16,
  "PIT", 5.36, 5.16,  28,
  "SDP", 3.27, 3.73,  22,
  "SEA", 3.61, 3.89,  19,
  "SFG", 2.81,    3,  20,
  "STL", 3.05, 3.66,  21,
  "TBR", 3.43, 3.31,  14,
  "TEX", 4.55, 3.94,  12,
  "TOR", 2.84, 3.69,  33,
  "WSN", 4.02, 4.43,  17
  )

top_12 <- tibble::tribble(
  ~playerid, ~ERA, ~FIP,
      16137, 2.37, 2.65,
      19361, 2.43, 1.63,
       3137, 2.46, 2.97,
      19374, 2.47, 3.16,
      16162, 2.56, 2.96,
      12703, 2.59, 4.03,
      22286, 2.64, 2.55,
       2520, 2.69, 3.32,
      10310, 2.78, 2.59,
      14107, 2.81,    3,
      11486, 2.84, 3.69,
      18679, 2.85, 3.14
  )

```

## Installation

The easiest way to get mlbplotR is to install it from  [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("mlbplotR")
```

To get a bug fix or use a feature from the development version, you can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("camdenk/mlbplotR")
```

## Examples

Let's plot every team on a grid with some extra customization:

```{r every-team, message = FALSE}
library(mlbplotR)
library(ggplot2)
library(dplyr)

teams_colors_logos <- mlbplotR::load_mlb_teams() |> 
  dplyr::filter(!team_abbr %in% c("AL", "NL", "MLB")) |> 
  dplyr::mutate(
    a = rep(1:6, 5),
    b = sort(rep(1:5, 6), decreasing = TRUE),
    # Keep alpha == 1 for teams that have an "A"
    alpha = ifelse(grepl("A", team_abbr), 1, 0.75),
    # Set teams that have an "E" to black & white
    color = ifelse(grepl("E", team_abbr), "b/w", NA) 
  )


 ggplot2::ggplot(teams_colors_logos, aes(x = a, y = b)) +
   mlbplotR::geom_mlb_logos(aes(team_abbr = team_abbr, 
                                color = color, alpha = alpha), 
                            width = 0.075) +
   ggplot2::geom_label(aes(label = team_abbr), nudge_y = -0.35, alpha = 0.5) +
   ggplot2::scale_color_identity() +
   ggplot2::scale_alpha_identity() +
   ggplot2::theme_void() 
```

This is a basic example with [FanGraphs](https://fangraphs.com) data (pulled using baseballr) which compares ERA to FIP for each team's top starting pitcher by ERA:

```{r scatter-example, message = FALSE}
# Uncomment this to pull data, currently commented out to save 
# a dependency on baseballr
# 
# library(baseballr)
# 
# df <- baseballr::fg_pitcher_leaders(x = 2021, y = 2021, 
#                                     q = 100, pitcher_type = "sta")
# 
# filtered_df <- df |>
#   dplyr::filter(Team != "- - -") |> 
#   # FanGraphs Team abbreviations aren't the standard
#   # for what's used in the plotting functions so 
#   # you could clean them with the following line, 
#   # but the geom_*_logos() functions should all
#   # clean the abbreviations before plotting
#   # 
#   # dplyr::mutate(Team = clean_team_abbrs(Team)) |> 
#   dplyr::group_by(Team) |> 
#   dplyr::slice_min(ERA, n = 1) |> 
#   dplyr::ungroup()

library(scales)

filtered_df |> 
  ggplot2::ggplot(aes(x = ERA, y = FIP)) +
  mlbplotR::geom_mlb_logos(aes(team_abbr = Team), width = 0.075, alpha = 0.7) +
  ggplot2::labs(title = "2021: ERA vs. FIP",
                subtitle = "Each Team's Top Starter By ERA | Min. 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold")) +
  ggplot2::scale_x_reverse(breaks = scales::pretty_breaks(), 
                           labels = scales::number_format(accuracy = 0.01),
                           expand = c(.1, .1)) +
  ggplot2::scale_y_reverse(breaks = scales::pretty_breaks(), 
                           labels = scales::number_format(accuracy = 0.01),
                           expand = c(.1, .1))
```

Here's another that looks at Home Runs Allowed by top pitchers:

```{r bar-example, message = FALSE}
filtered_df |> 
  # The scale_*_mlb() functions don't auto-clean abbreviations
  dplyr::mutate(Team = clean_team_abbrs(Team)) |> 
  ggplot2::ggplot(aes(x = Team, y = HR)) +
  ggplot2::geom_col(aes(color = Team, fill = Team), width = 0.5) +
  mlbplotR::geom_mlb_logos(aes(team_abbr = Team), width = 0.07, alpha = 0.9) +
  mlbplotR::scale_color_mlb(type = "secondary") +
  mlbplotR::scale_fill_mlb(alpha = 0.4) +
  ggplot2::labs(title = "2021: Home Runs Allowed For Top Pitchers",
                subtitle = "HRs Allowed By Each Team's Top Starter By ERA | Min. 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 axis.title.x = ggplot2::element_blank(),
                 axis.text.x = ggplot2::element_blank(),
                 panel.grid.major.x = element_blank()) +
  ggplot2::scale_x_discrete(expand = c(0.05, 0.075))
```

Instead of putting the logos in the plot, we can have them be the axis labels:

```{r axis-example, message = FALSE}
filtered_df |> 
  # The scale_*_mlb() functions don't auto-clean abbreviations
  dplyr::mutate(Team = clean_team_abbrs(Team)) |> 
  ggplot2::ggplot(aes(x = Team, y = HR)) +
  ggplot2::geom_col(aes(color = Team, fill = Team), width = 0.5) +
  #mlbplotR::geom_mlb_logos(aes(team_abbr = Team), width = 0.07, alpha = 0.9) +
  mlbplotR::scale_color_mlb(type = "secondary") +
  mlbplotR::scale_fill_mlb(alpha = 0.4) +
  ggplot2::labs(title = "2021: Home Runs Allowed",
                subtitle = "HRs Allowed By Each Team's Top Starter By ERA | Min. 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 axis.title.x = ggplot2::element_blank(),
                 panel.grid.major.x = ggplot2::element_blank(),
                 # this line triggers the replacement of team abbreviations with logos
                 axis.text.x = mlbplotR::element_mlb_logo()) +
  ggplot2::scale_x_discrete(expand = c(0.05, 0.075)) +
  ggplot2::scale_y_continuous(expand = c(0, 0, 0.05, 0))
```

The family of `element_*()` functions allows for a lot of extra customization with axes labels and can even be used for `facet_wrap()/facet_grid()` strip headers.



Lastly, here's an example using headshots:

```{r headshot-example, message = FALSE}
# top_12 <- df |>
#   dplyr::mutate(playerid = as.double(playerid)) |> 
#   dplyr::slice_min(ERA, n = 12)

top_12 |> 
  dplyr::left_join(mlbplotR::load_headshots(), 
                   by = c("playerid" = "fangraphs_id")) |> 
  ggplot(aes(x = ERA, y = FIP)) +
  mlbplotR::geom_mlb_headshots(aes(player_id = savant_id), height = 0.15) +
  ggplot2::scale_x_reverse(breaks = scales::pretty_breaks(), 
                           labels = scales::number_format(accuracy = 0.01),
                           expand = c(.1, .1)) +
  ggplot2::scale_y_reverse(breaks = scales::pretty_breaks(), 
                           labels = scales::number_format(accuracy = 0.01),
                           expand = c(.1, .1)) +
  ggplot2::labs(title = "2021 SP ERA Leaders",
                subtitle = "Minimum 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))

```

Note: If a player's headshot data can't be found, their headshot will be replaced with the MLB logo by default.
