---
title: "Getting Started with mlbplotR"
description: "A brief introduction to the ggplot2 extension."
author: "Camden Kay"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mlbplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 600
)
```

## Installation

mlbplotR is currently not on [CRAN](https://CRAN.R-project.org)
but you can get the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("camdenk/mlbplotR")
```

## Examples

Let's plot every team on a grid with some extra customization:

```{r every-team, message = FALSE}
library(mlbplotR)
library(ggplot2)
library(dplyr)

teams_colors_logos <- mlbplotR::load_mlb_teams() |> 
  dplyr::filter(!team_abbr %in% c("AL", "NL", "MLB")) |> 
  dplyr::mutate(
    a = rep(1:6, 5),
    b = sort(rep(1:5, 6), decreasing = TRUE),
    alpha = ifelse(grepl("A", team_abbr), 1, 0.75), # Keep alpha == 1 for teams that have an "A"
    color = ifelse(grepl("E", team_abbr), "b/w", NA) # Set teams that have an "E" to black & white
  )


 ggplot2::ggplot(teams_colors_logos, aes(x = a, y = b)) +
   mlbplotR::geom_mlb_logos(aes(team_abbr = team_abbr, color = color, alpha = alpha), width = 0.075) +
   ggplot2::geom_label(aes(label = team_abbr), nudge_y = -0.35, alpha = 0.5) +
   ggplot2::scale_color_identity() +
   ggplot2::scale_alpha_identity() +
   ggplot2::theme_void() 
```

This is a basic example with [FanGraphs](https://fangraphs.com) data (pulled using baseballr) which compares ERA to FIP for each team's top starting pitcher by ERA:

```{r scatter-example, message = FALSE}
library(baseballr)
library(scales)

df <- baseballr::fg_pitcher_leaders(x = 2021, y = 2021, q = 100, pitcher_type = "sta")

filtered_df <- df |>
  dplyr::filter(Team != "- - -") |> 
  # FanGraphs Team abbreviations aren't the standard
  # for what's used in the plotting functions so 
  # you could clean them with the following line, 
  # but the geom_*_logos() functions should all
  # clean the abbreviations before plotting
  # 
  # dplyr::mutate(Team = clean_team_abbrs(Team)) |> 
  dplyr::group_by(Team) |> 
  dplyr::slice_min(ERA, n = 1) |> 
  dplyr::ungroup()

filtered_df |> 
  ggplot2::ggplot(aes(x = ERA, y = FIP)) +
  mlbplotR::geom_mlb_logos(aes(team_abbr = Team), width = 0.075, alpha = 0.7) +
  ggplot2::labs(title = "2021: ERA vs. FIP",
                subtitle = "Each Team's Top Starter By ERA | Min. 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold")) +
  ggplot2::scale_x_reverse(breaks = scales::pretty_breaks(), expand = c(.1, .1)) +
  ggplot2::scale_y_reverse(breaks = scales::pretty_breaks(), expand = c(.1, .1))
```

Here's another that looks at Home Runs Allowed by top pitchers:

```{r bar-example, message = FALSE}
filtered_df |> 
  # The scale_*_mlb() functions don't auto-clean abbreviations
  dplyr::mutate(Team = clean_team_abbrs(Team)) |> 
  ggplot2::ggplot(aes(x = Team, y = HR)) +
  ggplot2::geom_col(aes(color = Team, fill = Team), width = 0.5) +
  mlbplotR::geom_mlb_logos(aes(team_abbr = Team), width = 0.07, alpha = 0.9) +
  mlbplotR::scale_color_mlb(type = "secondary") +
  mlbplotR::scale_fill_mlb(alpha = 0.4) +
  ggplot2::labs(title = "2021: Home Runs Allowed For Top Pitchers",
                subtitle = "HRs Allowed By Each Team's Top Starter By ERA | Min. 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 axis.title.x = ggplot2::element_blank(),
                 axis.text.x = ggplot2::element_blank(),
                 panel.grid.major.x = element_blank()) +
  ggplot2::scale_x_discrete(expand = c(0.05, 0.075))
```

Instead of putting the logos in the plot, we can have them be the axis labels:

```{r axis-example, message = FALSE}

filtered_df |> 
  # The scale_*_mlb() functions don't auto-clean abbreviations
  dplyr::mutate(Team = clean_team_abbrs(Team)) |> 
  ggplot2::ggplot(aes(x = Team, y = HR)) +
  ggplot2::geom_col(aes(color = Team, fill = Team), width = 0.5) +
  #mlbplotR::geom_mlb_logos(aes(team_abbr = Team), width = 0.07, alpha = 0.9) +
  mlbplotR::scale_color_mlb(type = "secondary") +
  mlbplotR::scale_fill_mlb(alpha = 0.4) +
  ggplot2::labs(title = "2021: Home Runs Allowed",
                subtitle = "HRs Allowed By Each Team's Top Starter By ERA | Min. 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 axis.title.x = ggplot2::element_blank(),
                 panel.grid.major.x = ggplot2::element_blank(),
                 # this line triggers the replacement of team abbreviations with logos
                 axis.text.x = mlbplotR::element_mlb_logo()) +
  ggplot2::scale_x_discrete(expand = c(0.05, 0.075))
```

The family of `element_*()` functions allows for a lot of extra customization with axes labels and can even be used for `facet_wrap()/facet_grid()` strip headers.



Lastly, here's an example using headshots:

```{r headshot-example, message = FALSE}

df |> 
  dplyr::mutate(playerid = as.double(playerid)) |> 
  dplyr::left_join(mlbplotR::load_headshots(), by = c("playerid" = "fangraphs_id")) |> 
  dplyr::slice_min(ERA, n = 12) |> 
  ggplot(aes(x = ERA, y = FIP)) +
  mlbplotR::geom_mlb_headshots(aes(player_id = savant_id), height = 0.15) +
  ggplot2::scale_x_reverse(breaks = scales::pretty_breaks(), expand = c(.1, .1)) +
  ggplot2::scale_y_reverse(breaks = scales::pretty_breaks(), expand = c(.1, .1)) +
  ggplot2::labs(title = "2021 SP ERA Leaders",
                subtitle = "Minimum 100 IP",
                caption = "Data: FanGraphs via baseballr") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))

```

Note: If a player's headshot data can't be found, their headshot will be replaced with the MLB logo.
